\textbf{Analyzing Machupo virus-receptor binding by molecular dynamics simulations}

### 2014-01-05 paper.tex
Pausing point.

\documentclass[journal=jacsat,manuscript=article]{achemso}

\usepackage{mathptmx} 
\usepackage{graphicx}
\usepackage{lscape}
\usepackage{amsmath}
\usepackage{verbatim}
\renewcommand{\bottomfraction}{.9}
\renewcommand{\topfraction}{.9}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{.9}

\title[Predicting mutant binding]{Predicting virus-receptor mutant binding by molecular dynamics simulation}

\author{Austin G. Meyer}
\email{austin.meyer@utexas.edu}
\phone{+1 512 971 0123}
\affiliation{Section of Integrative Biology, Institute for Cellular and Molecular Biology, and Center for Computational Biology and Bioinformatics, The University of Texas at Austin, Austin, TX 78731, USA}
\alsoaffiliation{School of Medicine, Texas Tech University Health Sciences Center, Lubbock, TX, 79430, USA}
\alsoaffiliation{Department of Molecular Biosciences, Institute for Cellular and Molecular Biology, The University of Texas at Austin, Austin, TX 78731, USA}

\author{Sara L. Sawyer}

\author{Andrew D. Ellington}
\affiliation{Department of Molecular Biosciences, Institute for Cellular and Molecular Biology, The University of Texas at Austin, Austin, TX 78731, USA}

\author{Claus O. Wilke}
\affiliation{Section of Integrative Biology, Institute for Cellular and Molecular Biology, and Center for Computational Biology and Bioinformatics, The University of Texas at Austin, Austin, TX 78731, USA}

\begin{document}

\newpage

\begin{abstract}
Existing computational methods to predict protein--protein interaction affinity often perform poorly in important test cases. In particular, the effects of multiple mutations, non-alanine substitutions, and flexible loops are difficult to predict with available tools and protocols. We present here a new method to interrogate affinity differences resulting from mutations in a host-virus protein--protein interface. Our method is based on extensive non-equilibrium all-atom simulations: We computationally pull the machupo virus (MACV) spike glycoprotein (GP1) away from the human transferrin receptor (hTfR1) and estimate affinity using the maximum applied force during a pulling simulation and the area under the force-versus-distance curve. We find that these quantities can provide novel biophysical insight into the GP1/hTfR1 interaction. First, with no prior knowledge of the system we can differentiate among wild-type and mutant complexes. Second, although the static co-crystal structure shows two large hydrogen-bonding networks in the GP1/hTfR1 interface, our simulations indicate that one of them may not be important for tight binding. Third, one viral site known to be critical for infection may mark an important evolutionary suppressor site for infection-resistant hTfR1 mutants. Finally, our method provides an elegant framework to compare the effects of multiple mutations, individually and jointly, on protein--protein interactions.
\end{abstract}

@@

The computational prediction of mutational effects on protein--protein interactions remains a challenging problem. Several methods are available to perform an energy difference calculation from an experimentally determined co-crystal structure. For example, computational alanine-scanning can be performed rapidly, with relatively low computational cost \cite{Grant2011,Kortemme2004}. However, such methods calculate the relative contribution to overall energy for each amino acid in a protein. They do not probe the interface directly, but rather calculate the total energy with each substitution \cite{Grant2011,Kortemme2004}. Alternative approaches have been developed using machine learning, training coefficients in a weighted equation containing geometric and energetic parameters \cite{Vreven2011,Vreven2012,Bajaj2011,Hwang2010}. Unfortunately, such machine-learning approaches often suffer in novel applications, for which available training sets are small or non-existent. As such, these methods are poorly suited for most host-virus protein--protein systems. By contrast, first principles methods offer the chance to forgo training. However, the currently available methods such as free energy perturbation (FEP) and thermodynamic integration (TI) rely on a two state model (where one state may be wild-type and the other may be a mutant) that transitions directly from one three-dimensional model to another with no intermediate steps \cite{Gilson1997,Lu2004,Chodera2011,Gumbart2013}. As a result, the accuracy of the measurement suffers as the two proteins become increasingly dissimilar; therefore, these techniques are largely limited to point mutant comparisons.

Here we propose a new method to probe mutational effects in protein--protein interactions directly based on first principles simulation, by pulling two proteins apart that are initially bound in a complex. We impart a pulling force within an all-atom molecular dynamics simulation on one member of the complex while the other is held in place. Then, we measure the force required for dissociation \cite{Gumbart2012,Lu1999,Park2004,Is2001A,Is2001B}. As a proxy for relative binding affinity, we use the maximum applied force required for separation and the area under the force-versus-distance curve (AUC).

We applied our method to interrogate the interaction between machupo virus (MACV) spike glycoprotein (GP1) and the human transferrin receptor (hTfR1) \cite{Abraham2010,Charrel2003}. Machupo is an ambisense RNA virus of the arenavirus family \cite{Charrel2003}. Worldwide, arenaviruses represent a significant source of emerging zoonotic diseases for the human population \cite{Charrel2003}. Members of the arenavirus family include the Lassa fever virus endemic to West Africa, the lymphochoriomeningitis virus (LCMV) endemic to rodents in several areas of the United States, and the Guanarito, Junin, and Machupo viruses endemic to rodents in South America \cite{Charrel2003}. The South American arenaviruses typically infect humans after rodent contamination and can cause a devastating hemorrhagic fever with high mortality \cite{Charrel2003}.

The hTfR1 is the primary receptor used by MACV for binding its host cell prior to infection. The primary role of hTfR1 $in~vivo$ is to bind transferrin for cellular iron uptake. The hTfR1 protein contains three extracellular domains: two basilar domains and an apical domain. The two basilar domains serve most of the transferrin-binding function \cite{Abraham2010,Rad20112}. Viral entry is initiated by GP1 binding to the apical domain of hTfR1. Previous work has indicated that the GP1/hTfR1 binding interaction is the primary determinant of MACV host range variation \cite{Rad20111,Rad20112}. The co-crystal structure shows the high affinity interaction between GP1 and hTfR1 forces the normally flexible loop in the apical domain of hTfR1 into a rigid beta-pleated sheet domain. For GP1, several extended loops mediate binding to hTfR1 \cite{Abraham2010,Rad20112}, and many of the interface interactions are mediated by extensive hydrogen-bonding networks \cite{Abraham2010}. Experimental alanine-scanning and whole-cell infectivity assays have identified several sites in both GP1 and hTfR1 that are probably critical for establishing infection \cite{Rad20111,Rad20112}.

We applied our computational method to wild type (WT) and mutant complexes, and found that we could resolve relative differences in unbinding and predict significant affinity changes. At sites known to be important for successful viral entry, we found that the biochemical cause of reduced infectivity may not be as simple as the static structure suggests. For example, the static structure shows a hydrogen-bonding network connected to site Asn348 in hTfR1. According to our simulations, this network may not affect binding affinity directly. In addition, our study offers an all-atom steered molecular dynamic approach to avoid the pitfalls of several existing methods used to evaluate mutations in protein--protein interfaces.

@@

We used the protein visualization software PyMOL \cite{PyMOL} to remove residues 121-190, 301-329, and 383-756 in the hTfR1. No residues were removed from the viral protein. We also removed 10 sugar residues from the native structure.

After system reduction, the Visual Molecular Dynamics (VMD) \cite{Humphrey1996} package along with its system of back-ends was used for all subsequent modeling. The Orient add-on package allowed us to rotate the system axis such that the direction of steering was oriented directly down the z-axis. De-glycosylation simplified the system such that Autopsf could easily find the chain terminations and patch them appropriately. The Solvate package was used to generate a TIP3P water model with a 5 \AA ngstrom buffer (relative to the maximum dimensions of the proteins) on all sides except down the positive z-axis where a 20 \AA ngstrom buffer was created. Finally, we used the Autoionize package to place 150 millimolar NaCl and neutralize the total system charge. In the end, each modeled system had approximately 28,000 atoms.

@@

NAMD was used for all simulations in this study \cite{Phillips2005}. In addition to the modeled system, for equilibration we generated a configuration file that fixed the $\alpha$-carbon backbone. This was accomplished by setting $\beta = 1$. Further, we generated a configuration file with fixed $\alpha$-carbon atoms at residues 41-92 (numbered linearly, in this case, starting at 1 for the first amino acid as was required for VMD) in the hTfR1. The second file was used to affix a harmonic restraint, thus preventing any unfolding due to system reduction. More importantly, the harmonic restraint allowed the protein complex to equilibrate while preventing any drift from its predefined orientation. Finally, we calculated the system center and dimensions for use in molecular dynamics settings.

We used the Charmm27 \cite{Brooks1983} all-atom force field. The initial system temperature was set to 310K. Several typical MD settings were used including switching and cutoff. In addition, we used a 2 femtosecond time step with rigid bonds (the exact configuration is provided in the supplement). We used periodic boundary conditions with the particle mesh ewald (PME) method of computing full system electrostatics outside of the explicit box. Furthermore, we used a group pressure cell, flexible box, and langevin piston during equilibration. Finally, a harmonic restraint (called harmonic constraint in VMD) was set as stated previously.

To start the simulation, the langevin piston was switched off and the system was minimized for 1000 steps. Next, the fixed backbone was released, and the system was minimized for an additional 1000 time steps. Subsequently, the system was released into all-atom molecular dynamics for 3000 steps. Finally, the langevin piston was turned on and the system was simulated for 2 ns (1,000,000 steps) of chemical time. For each mutant, twenty independent replicates were run with an identical protocol.

@@

Following equilibration, the final state of each simulation was used to generate a configuration file fixing the $\alpha$-carbon on residues 1, 58, 73-83, 96, 136, 137, 138, and 161 (again with linear numbering) in the hTfR1. Also, the $\alpha$-carbon of all residues (163-318 in linear numbering) in the GP1 were given an occupancy of 1, indicating they had an applied force during the simulation.

In every way possible the subsequent steering simulation had the same parameters as the equilibration simulation. Thus, the steered molecular dynamics (SMD) \cite{Is2001A,Is2001B} simulation was a typical restart with slightly different settings. Periodic boundary conditions were incorporated as part of the system and PME was again used to approximate full system electrostatics. In addition, we first tuned the pulling velocity for practical time considerations. Subsequently, the spring constant was adjusted to ensure a sufficient signal-to-noise ratio with the given velocity. As a result, the spring constant was set to 5~kcal/mol/\AA$^2$, velocity set to 1~\AA/ns, and direction down the positive $z$-axis.

Finally, SMD was run for 15 ns (7,500,000 time steps) of chemical time. For each simulation, we randomly selected one of the equilibration runs for restart. We ran 50 replicate simulations per mutant for a total of 550 SMD simulations. All GP1/hTfR1 complexes separated by greater than 4 \AA\ and many separated to 10 or more.

@@

@@

The python packages MDAnalysis \cite{Agrawal2011} and ProDy \cite{Bakan2011} were both used at various points in post-processing. The molecular trajectory (comprising the atomic coordinates per time) was parsed to compute the center-of-mass for each of the two complexes. The starting center-of-mass distance was set to zero and the distance was re-computed at each time step relative to the starting distance. The force was output directly to the standard log file.

The statistical package R was used for all further analysis and visualization. Each of the 50 independent trajectories per mutant produced a fairly noisy force curve. To reduce the noise, we averaged each point with a 50 point interpolation window. The window ensured that the averaged maximum applied force was never cut off near the beginning of the simulation due to lost time steps (caused by averaging). The two primary descriptive statistics we used were maximum interpolated applied force and total area under the interpolated curve (AUC). We used the pairwise.t.test function with the FDR p-value correction to test for significance. The ggplot \cite{ggplot} package was used to generate most of the figures.

@@

\subsection{The GP1/hTfR1 system}
The GP1/hTfR1 interface marks a particularly important and useful test system. There are several sites on both the human and viral protein known to affect the infectivity phenotype of MACV. Virtually all of the important sites have been mapped by \textit{in vitro} flow-cytometry based entry assays. The GP1/hTfR1 interface appears not to be dominated by one particular type of interaction (electrostatics, hydrogen-bonding, or van der Waals). In addition, much of the binding domain on hTfR1 is on a flexible loop and undergoes a large structural rearrangement upon binding, making many other computational techniques \cite{Grant2011,Kortemme2004} only marginally useful. The complex nature of this interface represents a particularly difficult challenge for traditional computational analysis.

For our experiments, we used the experimentally determined GP1/hTfR1 structure (PDBID: 3KAS) \cite{Abraham2010}. The apical domain of hTfR1 interacts directly with GP1 while the other two domains are closer to the cell membrane and have essentially no interaction with GP1. The biophysical independence of the apical domain allowed us to isolate it without significantly affecting the GP1/hTfR1 interaction. Figure~\ref{fig:protein_comparison} shows a model of the initial structure and that of the pared structure. Although GP1 has several glycosylatable residues, we opted to use the de-glycosylated protein for this study. The complexity of correctly parameterizing diverse sugar moieties is outside of the scope of this paper. Furthermore, although it is known that GP1 is glycosylated, and some of those sugars contact hTfR1, the sugars in the available PDB structure are not physiological for mammals \cite{Abraham2010}.

In total, we tested 7 point mutants and 3 double mutants in addition to the WT complex (Table~\ref{tab:summary_mutants} and Figure~\ref{fig:mutant_comparison}). Mutations at site hTfR1 site 211 have proven capable of causing loss-of-entry according to \textit{in vitro} flow-cytometry infection assays or known host-range limitations \cite{Rad2008,Rad20111,Rad20112}. Most likely, this effect is caused by the destruction of a critical hydrogen bond to Ser113 or Ser111 in GP1. The lost hydrogen bond would lead to the subsequent loss of a large hydrogen-bonding network seen in the crystal structure (Table~\ref{tab:summary_mutants}) \cite{Abraham2010}. In a manner similar to site 211, Asn348 appears to be important for binding by participating in a critical hydrogen bonding network \cite{Rad2008,Abraham2010} to GP1. In particular, Asn348Lys is reported in the literature to cause significantly reduced viral entry \textit{in vivo} (Table~\ref{tab:summary_mutants}) \cite{Rad2008,Abraham2010}. Finally, an alanine mutation at site 111 in GP1 (mutation vArg111Ala) has also been shown to cause decreased entry (Table~\ref{tab:summary_mutants}) \cite{Rad20112}. For notation purposes, the viral site is always referred to with a preceding `v'.

\subsection{Steered molecular dynamics simulations}

We proceeded to analyze this system using molecular dynamics. To ensure a natural solution structure, we performed an initial equilibration phase to reduce the intrinsic crystallization distortions. First, we generated a physically realistic simulation world around the protein complex. The environment included an explicitly modeled bath of water, and a physiologically relevant concentration of ions. Next, we carried out successive rounds of energy minimization with a fixed $\alpha$-carbon backbone, followed by heating, and backbone release. Last, we performed a short round of all-atom molecular dynamics with minimal restraints \cite{Cuendet2008}. We used the final state from each equilibrated system to restart another MD simulation. For SMD, at the centroid of several atoms a force was applied to a single point in GP1. By using a known spring constant, and pulling with a constant velocity, we could calculate the applied force during protein--protein dissociation \cite{Cuendet2008,Cuendet2011}. A typical averaged force curve comparison can be seen in Figure~\ref{fig:force_curve}. The force curves for each mutant were smoothed either by averaging over many replicates as in Figure~\ref{fig:force_curve} or by selecting an interpolation window for averaging. The latter approach was used for statistical analysis. In addition, to minimize the computational time in unproductive molecular dynamics, we shortened the simulation time such that the interaction just barely dropped to zero near the end of each run. As seen in Figure~\ref{fig:force_curve}, this distance was relatively consistent among mutants. The supplementary movie visually illustrates the separation distance between peptide domains. The quantities maximum applied force and AUC were derived from the force versus distances curves. Their summary statistics are reported in Table~\ref{tab:summary_statistics}. As we are more interested in the phenotypic impact of interface mutations we avoided many of the more physically rigorous, but technically complicated calculations that are possible with SMD \cite{Is2001A,Is2001B}.

Before systematically applying SMD to the GP1/hTfR1 interaction, we needed to ensure the method was sufficiently sensitive to distinguish between relatively minor point mutations. While SMD has been applied previously to measure the binding energy of high-affinity T-cell receptor interactions \cite{Cuendet2008,Cuendet2011}, to our knowledge the method has never been used previously to parse small differences in a complex energy landscape. For this initial sensitivity analysis, we tested alanine substitutions congruent with the traditional experimental and computational approach.

@@

This work was supported by the Defense Threat Reduction Agency (HDTRA1-12-C-0007) to A.D.E., S.L.S., and C.O.W., the National Science Foundation (MCB-0943383) and the Welch Foundation (F-1654) to A.D.E., the National Institutes of Health (R01-GM088344) to C.O.W., and the National Institutes of Health (R01-GM093086) to S.L.S.

\bibliography{paper}

>>>>

\documentclass[12pt]{article}

\usepackage[margin=2.5cm]{geometry}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{lineno}

\usepackage{times}
%correct punctuation for MBE
\bibpunct[,]{(}{)}{;}{a}{}{,}

\usepackage{amsmath}

\renewcommand{\bottomfraction}{.9}
\renewcommand{\topfraction}{.9}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{.9}

\linespread{1.5}
\begin{document}

\title{\textbf{Analyzing Machupo virus-receptor binding by molecular dynamics simulations}}
\author{Austin G. Meyer$^{1,2,3*}$, Sara L. Sawyer$^{2}$, Andrew D. Ellington$^{2}$,\\and Claus O. Wilke$^{1}$}
\date{}

\maketitle
\noindent
Address:\\
$^1$Department of Integrative Biology, Institute for Cellular and Molecular
Biology, and Center for\\Computational Biology and Bioinformatics. The University
of Texas at Austin, Austin, TX 78712, USA.\\
$^2$Department of Molecular Biosciences, Institute for Cellular and Molecular 
Biology, The University of Texas at Austin, Austin, TX 78712, USA.\\
$^3$School of Medicine, Texas Tech University Health Sciences Center, 
Lubbock, TX 79430, USA.\\

\bigskip
\noindent
$^*$Corresponding author\\
$\phantom{^*}$Email: austin.meyer@utexas.edu\\
$\phantom{^*}$Phone: +1 512 971 0123\\

\bigskip
\noindent
Manuscript type: research article

\bigskip
\noindent Keywords: protein--protein interaction, molecular dynamics, arenavirus, machupo

\linenumbers
\newpage

@@

The computational prediction of mutational effects on protein--protein interactions remains a challenging problem. Several methods are available to perform an energy difference calculation from an experimentally determined co-crystal structure. For example, end point methods can be performed rapidly, with relatively low computational cost \citep{Grant2011,Kortemme2004}. However, such methods can suffer from various simplifying assumptions. For example, they generally use an implicit solvent approximation and assume the end state difference with minimal structural rearrangement is sufficient to discriminate energetic differences \citep{Grant2011,Kortemme2004}. Alternative approaches have been developed using machine learning, training coefficients in a weighted equation containing geometric and energetic parameters \citep{Vreven2011,Vreven2012,Bajaj2011,Hwang2010}. Unfortunately, such machine-learning approaches often suffer in novel applications, for which available training sets are small or non-existent. As such, these methods are poorly suited for most host-virus protein--protein systems. By contrast, first principles methods can forgo training, but currently available methods such as free energy perturbation (FEP) and thermodynamic integration (TI) rely on a transitional model (where one state may be wild-type and the other may be a mutant) to make rigorous free energy calculations  \citep{Gilson1997,Lu2004,Chodera2011,Gumbart2013}. While these may be considered two of the gold standard techniques for calculating affinity differences, there are a huge number theoretical and technical complexities that must all be properly managed to ensure a converged solution \citep{Gumbart2013b}. Such considerations quickly come to dominate the protocol, and the necessary book keeping introduces the possibility of human error \citep{Gumbart2013b}. Moreover, as the two ending states look ever more dissimilar the chances of convergence falls rapidly. To ensure convergence, these techniques are typically limited to small differences (such as point mutant comparisons) with a few, very impressive exceptions \citep{Wang2006,Gumbart2013,Gumbart2013b}. For most, larger differences quickly become intractable as the number of intermediate steps required to compute a converged solution grows or the complexity of adding restraining potentials and computing approximations expands \citep{Wang2006,Gumbart2013,Gumbart2013b}.

Here we propose applying an existing method to a new test case to probe mutational effects in protein--protein interactions based on first principles simulation. We impart a pulling force within an all-atom molecular dynamics simulation on one member of the complex while the other is held in place. Then, we measure the force required for dissociation \citep{Gumbart2012,Lu1999,Park2004,Is2001A,Is2001B}. As a proxy for relative binding affinity, we use the maximum applied force required for separation and the area under the force-versus-distance curve (AUC). Although such biasing techniques are common in protein-ligand binding problems, they are less commonly applied to protein--protein interactions, and almost never to mutational analysis in a protein--protein system. This is largely the result of free energy convergence difficulties and computational limitations \citep{Cuendet2008,Cuendet2011}; using maximum force or simple AUC can solve these problems. We also used the traditional dual topology FEP technique to compute relative free energy differences for comparison to our simplified force biased approach.

We used SMD and FEP to interrogate the interaction between machupo virus (MACV) spike glycoprotein (GP1) and the human transferrin receptor (hTfR1) \citep{Abraham2010,Charrel2003}. Machupo virus is an ambisense RNA virus of the arenavirus family \citep{Charrel2003}. Worldwide, arenaviruses represent a significant source of emerging zoonotic diseases for the human population \citep{Charrel2003}. Members of the arenavirus family include the Lassa fever virus endemic to West Africa, the lymphochoriomeningitis virus (LCMV) endemic to rodents in several areas of the United States, and the Guanarito, Junin, and Machupo viruses endemic to rodents in South America \citep{Charrel2003}. The South American arenaviruses typically infect humans after rodent contamination and can cause a devastating hemorrhagic fever with high mortality \citep{Charrel2003}.

The hTfR1 is the primary receptor used by MACV for binding its host cell prior to infection. The primary role of hTfR1 $in~vivo$ is to bind transferrin for cellular iron uptake. The hTfR1 protein contains three extracellular domains: two basilar domains and an apical domain. The two basilar domains serve most of the transferrin-binding function \citep{Abraham2010,Rad20112}. Viral entry is initiated by GP1 binding to the apical domain of hTfR1. Previous work has indicated that the GP1/hTfR1 binding interaction is the primary determinant of MACV host range variation \citep{Rad20111,Rad20112}. The co-crystal structure shows the high affinity interaction between GP1 and hTfR1 forces the normally flexible loop in the apical domain of hTfR1 into a rigid $\beta$-pleated sheet domain. For GP1, several extended loops mediate binding to hTfR1 \citep{Abraham2010,Rad20112}, and many of the interface interactions are mediated by extensive hydrogen-bonding networks \citep{Abraham2010}. Experimental alanine-scanning and whole-cell infectivity assays have identified several sites in both GP1 and hTfR1 that are probably critical for establishing infection \citep{Rad20111,Rad20112}.

We applied our computational method to wild type (WT) and mutant complexes, and found that we could resolve relative differences in unbinding and predict significant affinity changes. Importantly, the affinity changes predicted using only max force or AUC show a strong correlation with rigorous relative free energy differences computed by FEP. At sites known to be important for successful viral entry, we found that the biochemical cause of reduced infectivity may not be as simple as the static structure suggests. For example, the static structure shows a hydrogen-bonding network connected to site Asn348 in hTfR1. According to our simulations, this network may not affect binding affinity directly. In addition, our study offers an all-atom steered molecular dynamic approach to avoid some of the pitfalls of several existing methods used to evaluate mutations in protein--protein interfaces.

@@

For our experiments, we used the experimentally determined GP1/hTfR1 structure (PDBID: 3KAS) \citep{Abraham2010}. The apical domain of hTfR1 interacts directly with GP1 while the other two domains are closer to the cell membrane and have essentially no interaction with GP1. The biophysical independence of the apical domain allowed us to isolate it without significantly affecting the GP1/hTfR1 interaction.

We used the protein visualization software PyMOL \citep{PyMOL} to remove residues 121-190, 301-329, and 383-756 in the hTfR1. No residues were removed from the viral protein.  Figure~\ref{fig:protein_comparison} shows a model of the initial structure and that of the pared structure. Although GP1 has several glycosylatable residues, we opted to use the de-glycosylated protein for this study. The complexity of correctly parameterizing diverse sugar moieties is outside of the scope of this paper. Furthermore, although it is known that GP1 is glycosylated, and some of those sugars contact hTfR1, the sugars in the available PDB structure are not physiological for mammals \citep{Abraham2010}.  In total we removed 10 sugars from the crystal structure for this study.

After system reduction, the Visual Molecular Dynamics (VMD) \citep{Humphrey1996} package along with its system of back-ends was used for all subsequent modeling. The Orient add-on package allowed us to rotate the system axis such that the direction of steering was oriented directly down the z-axis. De-glycosylation simplified the system such that Autopsf could easily find the chain terminations and patch them appropriately. The Solvate package was used to generate a TIP3P water model with a 5 \AA ngstrom buffer (relative to the maximum dimensions of the proteins) on all sides except down the positive z-axis where a 20 \AA ngstrom buffer was created. Finally, we used the Autoionize package to place 150 millimolar NaCl and neutralize the total system charge. In the end, each modeled system had approximately 28,000 atoms.

@@

NAMD was used for all simulations in this study \citep{Phillips2005}. In addition to the modeled system, for equilibration we generated a configuration file that fixed the $\alpha$-carbon backbone. This was accomplished by setting the B-factor column to 1 for the fixed atoms and to zero for all other atoms. Further, we generated a configuration file with fixed $\alpha$-carbon atoms at residues 41-92 (numbered linearly, in this case, starting at 1 for the first amino acid as was required for NAMD) in the hTfR1. The second file was used to affix a harmonic restraint, thus preventing any unfolding due to system reduction. More importantly, the harmonic restraint allowed the protein complex to equilibrate while preventing any drift from its predefined position; the restraint did not constrain the structure of each protein, or the relative position or orientation of the two proteins to each other. Finally, we calculated the system center and dimensions for use in molecular dynamics settings.

The exact NAMD configuration file is made available as a supplement and a github repository with all necessary set up files is provided. We used the Charmm27 \citep{Brooks1983} all-atom force field. The initial system temperature was set to 310K. Several typical MD settings were used including switching and cutoff distances (see provided configuration file). In addition, we used a 2 femtosecond time step with rigid bonds. We used periodic boundary conditions with the particle mesh ewald (PME) method of computing full system electrostatics outside of the explicit box. Furthermore, we used a group pressure cell, flexible box, langevin barostat and lavegin thermostat during equilibration. A harmonic restraint (called harmonic constraint in VMD) was set as stated previously.

To start the simulation, the barostat was switched off and the system was minimized for 1000 steps. Next, the fixed backbone was released, and the system was minimized for an additional 1000 time steps. Subsequently, the system was released into all-atom molecular dynamics for 3000 steps. Finally, the langevin piston was turned on and the system was simulated for 2 ns (1,000,000 steps) of chemical time. For each mutant, twenty independent equilibration replicates were run with an identical protocol.

@@

We used the final state from each equilibrated system to restart another MD simulation. Our steering protocol is fundamentally similar to \citet{Cuendet2008} with slightly different parameter choices. Perhaps the one significant difference lies in our choosing to not use a thermostat or barostat. However, this should not be a problem since we are not trying to calculate the binding free energy by any physically rigorous approach (the Jarzynski inequality being one example). Following equilibration, the final state of each simulation was used to generate a configuration file fixing the $\alpha$-carbon on residues 1, 58, 73-83, 96, 136, 137, 138, and 161 (again with linear numbering) in the hTfR1. These residues were selected as they are far from the binding interface and sufficiently distributed to prevent any orientational motion of the receptor relative to the viral spike protein. The $\alpha$-carbon of all residues (163-318 in linear numbering) in GP1 received an applied force during the simulation. The NAMD convention does not actually apply a force to all $\alpha$-carbon atoms, but rather uses the selection to compute an initial center of mass. Then, during the steering run, the single center of mass point is pulled with the parameters desribed below. We used the same force field parameters (exclude, cutoff, switching, etc.), the same integrator parameters (time step, rigidbonds on, all molecular being wrapped, etc.), and the same particle mesh ewald parameters as in equilibration. Periodic boundary conditions were incorporated as part of the system (as is the convention in NAMD restart) and PME was again used to approximate full system electrostatics.

We ran test simulations at several force constants, and visually inspected the results. A force constant of 5~kcal/mol/\AA$^2$ was chosen due to its relatively low signal-to-noise ratio. This is slightly lower than the more common 7~kcal/mol/\AA$^2$ found in several recent studies; however, it seems that value is so commonly selected primarily because it is the force constant found in the NAMD, SMD tutorial available through the developers. Moreover, the force constant could very likely be set to a range of nearby values with little loss in predictive power. In SMD experiments the pulling velocity should be as low as possible for the available computational time \citet{Cuendet2008,Cuendet2011}. Thus we choose a velocity of 0.000001~\AA/fs = 1~\AA/ns, and direction down the positive $z$-axis. This number could be set to pull faster if the computing time must be reduced, but slower than necessary pulling speeds are not typically considered problematic.

SMD was run for 15 ns (7,500,000 time steps) of chemical time. For each simulation, we randomly selected one of the equilibration runs for restart. We ran 50 replicate simulations per mutant for a total of 550 SMD simulations. All GP1/hTfR1 complexes separated by greater than 4 \AA\ and many separated to 10 or more.

@@

\subsection{Free Energy Perturbation}

Briefly, we used the traditional dual topology approach to FEP. This involves a thermodynamic cycle where a set of atoms are progressively decoupled from the environment while another set of atoms are progressively coupled. To compute the relative free energy difference requires knowing the free energy change when the transformation is carried out for the bound complex and the individual protein. Then, one can compute the relative free energy difference between a WT and mutant complex by taking the difference between the energy required to decouple/couple the atoms in solution from the energy required to decouple/couple the atoms in the bound complex.

Again, the NAMD configuration file is made available as a supplemental file. We used a similar configuration to that in equilibration. One significant difference was to make a cubic water box with a side length equal to the long axis of the complex plus a 10~\AA buffer on either side, and simply restrict center of mass motion with the NAMD setting. This was done to avoid affecting the system energy while calculating free energy differences.

The transition protocol for bound and free protein systems were identical. They started with 1000 steps of minimization and 250,000 steps of equilibration in the starting state for the forward and reverse directions. Phase transitions were carried out in steps of $\lambda$=0.05. Each transition was carried out for 250,000 steps. The first 100,000 steps after phase transition were reserved for equilibration and the final 150,000 steps were used for data collection.

The FEP analysis tool in VMD was used for subsequent analysis. We used it to perform error analysis and compute the Bennett acceptance ratio as the maximum likelihood free energy difference of the two states under consideration. Though the larger transitions presented difficulty in a small number of windows, forward and reverse hysteresis was generally in good agreement for all complexes. The double mutants were performed by first doing the Y211A mutation followed by the other of the two mutants. Then, the numbers were simply added together to get the total energetic difference.

@@

The python packages MDAnalysis \citep{Agrawal2011} and ProDy \citep{Bakan2011} were both used at various points in post-processing. The molecular trajectory (comprising the atomic coordinates per time) was parsed to compute the center-of-mass for each of the two complexes. The starting center-of-mass distance was set to zero and the distance was re-computed at each time step relative to the starting distance.

The statistical package R was used for all further analysis and visualization. Each of the 50 independent trajectories per mutant produced a fairly noisy force curve. The force curves for each mutant were smoothed either by averaging over many replicates as in Figure~\ref{fig:force_curve} or by selecting a 50 frame interpolation window for averaging. The latter approach was used for statistical analysis. The two primary descriptive statistics we used were maximum interpolated applied force and total area under the interpolated curve (AUC). We used the pairwise.t.test function with the FDR p-value correction to test for significance. The ggplot \citep{ggplot} package was used to generate most of the figures.

@@

\subsection{The GP1/hTfR1 system}
The GP1/hTfR1 interface marks a particularly important and useful test system. There are several sites on both the human and viral protein known to affect the infectivity phenotype of MACV. Many of the important sites have been mapped by \textit{in vitro} flow-cytometry based entry assays. The GP1/hTfR1 interface appears not to be dominated by one particular type of interaction (electrostatics, hydrogen-bonding, or van der Waals). In addition, much of the binding domain on hTfR1 is on a loop that is flexible prior to viral binding, but organizes to become a strand of a $\beta$-sheet on binding. As a result,  many other computational techniques \citep{Grant2011,Kortemme2004} only marginally useful. The complex nature of this interface represents a particularly difficult challenge for traditional computational analysis.

In total, we tested 7 point mutants and 3 double mutants in addition to the WT complex (Table~\ref{tab:summary_mutants} and Figure~\ref{fig:mutant_comparison}). All of the mutations are within 5~\AA of the protein--protein interface. Mutations at site hTfR1 site 211 have proven capable of causing loss-of-entry according to \textit{in vitro} flow-cytometry infection assays or known host-range limitations \citep{Rad2008,Rad20111,Rad20112}. Most likely, this effect is caused by the destruction of a critical hydrogen bond to Ser113 or Ser111 in GP1. The lost hydrogen bond would lead to the subsequent loss of a large hydrogen-bonding network seen in the crystal structure (Table~\ref{tab:summary_mutants}) \citep{Abraham2010}. In a manner similar to site 211, Asn348 appears to be important for binding by participating in a critical hydrogen bonding network \citep{Rad2008,Abraham2010} to GP1. In particular, Asn348Lys is reported in the literature to cause significantly reduced viral entry \textit{in vivo} (Table~\ref{tab:summary_mutants}) \citep{Rad2008,Abraham2010}. Finally, an alanine mutation at site 111 in GP1 (mutation vArg111Ala) has also been shown to cause decreased entry (Table~\ref{tab:summary_mutants}) \citep{Rad20112}. For notation purposes, the viral site is always referred to with a preceding `v'.

Despite, the fact that viral binding occurs at the site of a flexible loop in the free hTfR structure, our data shows after binding the strand is extremely rigid. None of the sites in the interfacial strand have root mean squared fluctuation (RMSF) values in the top half of all sites in the complex during equilibration (figure). This is unsurprising considering the high degree of burial that occurs as a result of viral binding. In addition, computing the root mean squared deviation (RMSD) of the entire structure over the trajectory shows that none of the mutations are so deleterious as the cause rapid unbinding. In fact, the RMSD over trajectory looks incredibly invariant across mutants (figure).

\subsection{Molecular dynamics simulations}

We proceeded to analyze this system using two molecular dynamics techniques. First, in SMD by using a known spring constant, and pulling with a constant velocity, we could calculate the applied force during protein--protein dissociation \citep{Cuendet2008,Cuendet2011}. A typical averaged force curve comparison can be seen in Figure~\ref{fig:force_curve}. As seen in Figure~\ref{fig:force_curve}, this distance was relatively consistent among mutants. The supplementary movie visually illustrates the separation distance between peptide domains. The quantities maximum applied force and AUC were derived from the force versus distances curves. Their summary statistics are reported in Table~\ref{tab:summary_statistics}. As we are more interested in the phenotypic impact of interface mutations we avoided many of the more physically rigorous, but technically complicated calculations that are possible with SMD \citep{Is2001A,Is2001B}. Indeed, our set up would not have allowed rigorous calculations as we omitted several tools such as a barostat and a thermostat that would be required for an NPT ensemble.

Before systematically applying SMD to the GP1/hTfR1 interaction, we needed to ensure the method was sufficiently sensitive to distinguish between relatively minor point mutations. While SMD has been applied previously to measure the binding energy of high-affinity T-cell receptor interactions \citep{Cuendet2008,Cuendet2011}, to our knowledge the method has never been used previously to parse small differences in a complex energy landscape. For this initial sensitivity analysis, we tested alanine substitutions congruent with the traditional experimental and computational approach. We proceeded to compare our SMD results to that of the standard dual topology FEP approach to calculate relative free energy differences. The correlation between the energetically rigorous FEP and our statistical approach is high. For all 11 complexes tested, the correlation between max force and FEP was r=0.767 at p=0.00576, and the correlation between AUC and FEP was r=0.72 at p=0.01 (figure). As a result of the strong correlation, we will generally refer to the SMD differences only.

@@

\bibliographystyle{MBE}
\bibliography{paper}
